Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-29 main.tex
shrink to fit

% Configuration for minted package
\setminted{
    style=colorful, % Set code style to colorful
    fontsize=\scriptsize,
    linenos, % Enable line numbers
    breaklines, % Enable line breaks
    breakanywhere, % Allow line breaks anywhere
    autogobble,
}
\setmintedinline{fontsize=\small}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

>>>>

% Configuration for minted package
\setminted{
    style=colorful, % Set code style to colorful
    fontsize=\scriptsize,
    linenos=false, % Enable line numbers
    breaklines, % Enable line breaks
    breakanywhere, % Allow line breaks anywhere
    autogobble,
}
\setmintedinline{fontsize=\small}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

### 2024-03-29 sections/compiler.tex
shrink to fit

\subsection{Wrapperization}

    Many fancy operations on indices can be resolved by introducing equivalent
    \textbf{wrapper arrays} which modify the behavior of the tensors they wrap,
    or by introducing \textbf{mask arrays} which replace index expressions like
    \mintinline{julia}{i == j} with their equivalent masks (in this case, a
    diagonal mask tensor).

@@

    For example, \begin{minted}{julia}
    for i=_, j=_
        if i <= j
            s[] += A[i, j]
        end
    end
    \end{minted}

    would likely compile to something like

    \begin{minted}{julia}
    for i = 1:n
        for j = 1:i
            s[] += A[i, j]
        end
    end
    \end{minted}

@@

    In Finch, we use a straightforward algorithm to automatically determine suitable dimensions for loops and declaration statements (output tensors).
    %
    Finch determines the dimension of a loop index i from all of the tensors using i in an access, as well as the bounds in the loop itself, and operates similarly for declarations.
    %   
    Our algorithm uses the following principles, assuming dimension types form a lattice:
    \begin{enumerate}
        \item We assign dimensions to indices.
        \item Using an index in an access “hints” that the index should have the corresponding dimension.
        \item Loop dimensions are equal to the “meet” of all hints in the loop body
        and any existing dimensions in the loop bounds. The meet usually asserts
        that dimensions match, but may also e.g. propagate info about parallelization
        \item The \mintinline{julia}{_} symbol represents a dimensionless quantity at the bottom of the dimension lattice.
        \item We assign dimensions to declarations.
        \item Left hand side (updating) tensor access “hint” the size of that tensor
        \item The dimensions of a declaration are the “meet” of all hints from
        the declaration to the first read.
        \item The new dimensions of the declared tensor are used when the tensor is on the right hand side (reading)
        access.
    \end{enumerate}

    For example, consider the following code

    \begin{minted}{julia}
    A = fsprand(3, 4, 0.5)
    B = fsprand(4, 5, 0.5)
    C = Tensor(Dense(SparseList(Element(0.0))))
    @finch begin
    C .= 0
    for i = 1:3
        for j = _
            for k = _
                C[i, j] += A[i, k] * B[k, j]
            end
        end
    end
    \end{minted}

    In the above code, the second dimension of A must match the first dimension of B. Also, the first dimension of A must match the i loop dimension, 1:3. Finch will also resize declared tensors to match indices used in writes, so C is resized to (1:3, 1:5). If no dimensions are specified elsewhere, then Finch will use the dimension of the declared tensor.

    Dimensionalization occurs after wrapper arrays are de-sugared. You can
    therefore exempt a tensor from dimensionalization by wrapping the
    corresponding index in ~ to produce a "PermissiveArray". For example,

    \begin{minted}{julia}
    @finch begin
    y .= 0
    for i = 1:3
        y[~i] += x[i]
    end
    \end{minted}

    does not set the dimension of y, and y does not participate in dimensionalization.

@@

    We make expressions concordant by inserting loops with one iteration. For example, the following expression can be made concordant by adding a loop for the out of order access to \texttt{i}.
    \begin{minted}{julia}
        @finch begin
        for i = _
            for j = _
                for k = _
                    s[] += A[i, k, j]
                end
            end
        end
    \end{minted}

    \begin{minted}{julia}
        @finch begin
        for i = _
            for j = _
                for k = _
                    for l = _
                        if i == l #diagonal mask
                            s[] += A[l, k, j]
                        end
                    end
                end
            end
        end
    \end{minted}

    We can concordize unrecognized index expressions using a similar approach:

    \begin{minted}{julia}
        for i = _
            A[I[i]] += 1
        end
    \end{minted}

    \begin{minted}{julia}
        for i = _
            for j = _
                if j == I[i]
                    A[j] += 1
                end
            end
        end
    \end{minted}

\subsection{Life Cycles and Scopes}
\teo{Can we maybe skip this because we cover it in the semantics?}
\willow{we should at least say here that we have a pass that adds freeze and thaw appropriately}
\subsection{Simplification and Algebraic Transformations}

\subsection{Bounds Analysis}
\subsection{Performance Warnings}
\teo{I also think this maybe belongs in an appendix.}

>>>>

\subsection{Wrapperization}
    \begin{wrapfigure}{R}{0.2\textwidth}
        \vspace{-1\intextsep}
        \begin{minted}{julia}
            for i=_, j=_
                if i <= j
                    s[] += A[i, j]
                end
            end
        \end{minted}
        $\downarrow$
        \begin{minted}{julia}
            for i=_, j=_
                if UpTriMask()[i, j]
                    s[] += A[i, j]
                end
            end
        \end{minted}
        $\downarrow$
        \begin{minted}{julia}
            for i = 1:n
                for j = 1:i
                    s[] += A[i, j]
                end
            end
        \end{minted}
        \caption{Wrapperization}\label{fig:wrapperization}
    \end{wrapfigure}

    Many fancy operations on indices can be resolved by introducing equivalent
    \textbf{wrapper arrays} which modify the behavior of the tensors they wrap,
    or by introducing \textbf{mask arrays} which replace index expressions like
    \mintinline{julia}{i <= j} with their equivalent masks (in this case, a
    triangular mask tensor).

@@

@@

    In Finch, we use a straightforward algorithm to automatically determine suitable dimensions for loops and declaration statements (output tensors).
    %
    Finch determines the dimension of a loop index i from all of the tensors using i in an access, as well as the bounds in the loop itself, and operates similarly for declarations.
    %   
    Our algorithm uses the following principles, assuming dimension types form a lattice:
    \begin{wrapfigure}{r}{0.3\textwidth}
        \begin{minted}{julia}
            #A is 3 x 4
            #B is 4 x 5
            C .= 0
            for i = 1:3
                for j = _
                    for k = _
                        C[i, j] += A[i, k] * B[k, j]
                    end
                end
            end
        \end{minted}
        $\downarrow$
        \begin{minted}{julia}
            C .= 0
            for i = 1:3
                for j = 1:5
                    for k = 1:4
                        C[i, j] += A[i, k] * B[k, j]
                    end
                end
            end
        \end{minted}
        \caption{Dimensionalization}\label{fig:dimensionalization}
    \end{wrapfigure}
    \begin{enumerate}
        \item We assign dimensions to indices.
        \item Using an index in an access “hints” that the index should have the corresponding dimension.
        \item Loop dimensions are equal to the “meet” of all hints in the loop body
        and any existing dimensions in the loop bounds. The meet usually asserts
        that dimensions match, but may also e.g. propagate info about parallelization
        \item The \mintinline{julia}{_} symbol represents a dimensionless quantity at the bottom of the dimension lattice.
        \item We assign dimensions to declarations.
        \item Left hand side (updating) tensor access “hint” the size of that tensor
        \item The dimensions of a declaration are the “meet” of all hints from
        the declaration to the first read.
        \item The new dimensions of the declared tensor are used when the tensor is on the right hand side (reading)
        access.
    \end{enumerate}

    For example, in Figure ~\ref{fig:dimensionalization}, the second dimension of A must match the first dimension of B. Also, the first dimension of A must match the i loop dimension, 1:3. Finch will also resize declared tensors to match indices used in writes, so C is resized to (1:3, 1:5). If no dimensions are specified elsewhere, then Finch will use the dimension of the declared tensor.

    Dimensionalization occurs after wrapper arrays are de-sugared. You can
    therefore exempt a mode in an access from dimensionalization by wrapping the
    corresponding index in $~$ to produce a "PermissiveArray" (e.g. $A[~i]$).

@@

    We make expressions concordant by inserting loops with one iteration. For example, the following expression can be made concordant by adding a loop for the out of order access to \texttt{i}.
    \begin{figure}[htbp]
        \centering % Center the figure
        % First pair of examples
        \begin{minipage}{0.3\textwidth}
        \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
        for i = _
            for j = _
                for k = _
                    s[] += A[i, k, j]
                end
            end
        end
        \end{minted}
        \end{minipage}
        $\rightarrow$
        \begin{minipage}{0.4\textwidth}
        \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
            for i = _
                for j = _
                    for k = _
                        for l = _
                            if i == l
                                s[] += A[l, k, j]
                            end
                        end
                    end
                end
            end
        \end{minted}
        \end{minipage}\hfill%
        \begin{minipage}{0.2\textwidth}
            \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
                for i = _
                    A[I[i]] += 1
                end
                \end{minted}
            $\downarrow$
            \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
            for i = _
                for j = _
                    if j == I[i]
                        A[j] += 1
                    end
                end
            end
            \end{minted}
        \end{minipage}

        \caption{Transformations demonstrating concordization of unrecognized index expressions.}
        \end{figure}
\begin{wrapfigure}{r}{0.45\textwidth}
    \vspace{-1\intextsep}
    \begin{minipage}{0.2\textwidth}
    \begin{minted}{julia}
        y .= 0
        for i = _
            y[i] .= x[i] + 1
        end
        for i = _
            x[i] += 1
            y[i] += 1
        end 
        for i = _
            x[i] += y[i]
        end
    \end{minted}
    \end{minipage}
    $\rightarrow$
    \begin{minipage}{0.2\textwidth}
    \begin{minted}{julia}
        y .= 0
        for i = _
            y[i] .= x[i] + 1
        end
        @thaw(x)
        @freeze(y)
        for i = _
            x[i] += 1
            y[i] += 1
        end
        @freeze(x)
        for i = _
            x[i] += y[i]
        end
    \end{minted}
    \end{minipage}
    \caption{Life Cycle Automation.}\label{fig:lifecycles}
\end{wrapfigure}
\subsection{Life Cycle Automation}
Finally, we introduce a pass to avoid needed to manually insert the \mintinline{julia}{@freeze} or \mintinline{julia}{@thaw} macros. The pass will insert these statements at the appropriate place in the program if they have not been inserted already, easing the burden on the programmer and bridging between structured and dense languages. An example is given in Figure \ref{fig:lifecycles}.
